\section*{Results}\label{s:results}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Systematic colocalization of variants from genome-wide associations studies and expression quantitative trait loci}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

To annotate traits from genome-wide associations studies (GWAS) with genes and tissues,
I developed a pipeline that systematically colocalizes variants from GWAS and expression quantitative trait loci (eQTLs) (Section Methods).
%
127 eQTL studies were downloaded from the EBI eQTL catalogue, which aims to provide uniformly processed eQTLs in many
tissues and cell types \citep{2021.Alasoo.Kerimov} (Supplementary table 1).

GWAS traits were selected from the IEU OpenGWAS database based on four criteria \citep{2018.Parkinson.Buniello}.
%
The first criterion was to exclude molecular traits such as proteome or methylome.
%	
The second criterion was to include only the European population, because most
samples from the EBI eQTL catalogue belong to the European population.
%
The third criterion was to keep only well-defined medical or physiological
conditions and exclude environmental traits such "employment status" or "self-reported" medical conditions.
%
The fourth criterion was to keep only GWAS with at least 10,000 subjects, 2,000 controls and 2,000 cases.
%
These filters resulted in 418 GWAS traits (Supplementary table 2).
%
These studies resulted in 10,621 clumped leading variants with a p-value below 5e-8 from 335 GWAS (Supplementary file).
% select count(distinct gwas_id) from tophits; 335
% select count(*) from (select DISTINCT chrom, pos, rsid, ea from tophits) as sel; 10621
% tophits goes to OSF
% colocpleio goes to OSF

My pipeline uncovered eQTL colocalizations for 5,484 or 64$\%$ of the leading GWAS variants with 7,040 eQTL genes and 127 eQTL biosamples.
% select count(*) from (select distinct tophits.chrom,tophits.pos,tophits.ea,tophits.rsid from tophits inner join coloc
% on coloc.rsid=tophits.rsid where coloc.pp_h4_abf>=0.75) as sel; 5484
% select count(*) from (select distinct eqtl_gene_id from coloc where pp_h4_abf>=0.75) as sel;  7040
% select count(*) from (select distinct eqtl_id from coloc where pp_h4_abf>=0.75) as sel;  127



The largest classes of diseases by number of leading GWAS variants are cancer of breast (1976), allergy (1288), and autoimmune diseases (1216).
%
In these disease classes, the percentages of variants with a colocalized eQTL are 58\%, 61\% and 68\%, respectively (Supplementary table ST3).
% Supplementary ST3
% Number of tophits rsid per GWAS class
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits where tophits.gwas_id=gwas_annot.gwas_id group by gwas_class order by count desc;
% Number of tophits rsid per GWAS class that is also preset in coloc table
% select count(distinct tophits.rsid),gwas_class from gwas_annot, tophits, coloc where tophits.gwas_id=gwas_annot.gwas_id and tophits.rsid=coloc.rsid group by gwas_class order by count desc;

These results is much higher than previous colocalization analyses for autoimmune diseases by previous authors that
found 23-41\% depending on the study \citep{2016.Soranzo.Chen,2017.Cotsapas.Chun,2021.Li.Mu}.
% Discuss 25% in Chun et al 2017, Nature Genetics
% Discuss 41% (eQTL + sQTLs) in Mu et al 2021, Genome Biology
% Discuss 23% in Chen et al 2016, Cell. Genetic Drivers of Epigenetic and Transcriptional Variation in Human Immune Cells

These data is a useful resource to annotate GWAS variants with genes and biosamples.
Therefore I have developed a web application where I expose 138,301 variants from 293 GWAS that colocalize with at least
one eQTL at a threshold value PP.H4.abf$\geq$0.75.
% TODO
% select count(*) from (select DISTINCT chrom,pos38,alt FROM colocweb) as sel; 138301
% select count(*) from (select DISTINCT gwas_id FROM colocweb) as sel; 293

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% Work with PP.H4.abf`>=0.75 and SNP.PP.H4>=0.5
\subsection*{Traits classification based on the effect size of eQTLs}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In the previous colocalization analysis, we found many eQTLs that are associated to very different traits.
%
To systematically identify pleiotropic eQTLs, I manually annotated the GWAS traits in different classes.

Then I wanted to verify if different traits cluster coherently in different trait classes.
%
As trait distance, I used the the spearman correlation of the eQTL splitted by
chromosome, position, alternative allele, eQTL gene and eQTL biosample.
%
To avoid non resolve eQTLs, I set a cutoff SNP.PP.H4>=0.5 in addition to the cutoff PP.H4.abf>=0.75.
%
The cutoff SNP.PP.H4>=0.5 was chosen, it allow to group pleiotropic eQTLs in a reasonable number of regions (See below).
%
I plotted these trait distances as a hierarchically-clustered heatmap.
%
Only trait classes with more than a minimum number of traits were plotted (Fig. \ref{fig:gwas_distance}).
%
Due to the large number of datasets, I have create a separate class for some cancers such as breast and skin.
%
In the OpenGWAS, some traits are redundant because the come from the same reference.
%
I plotted non-rendundant GWAS traits.
%
This analysis shows a coherent clustering of traits belonging to autoimmune, circulatory system diseases, cancers and allergies (Fig. \ref{fig:gwas_distance}).
%
Breast and skin cancer are found in two different cluster with some shared signal.
%
Multiple sclerosis cluster with autoimmune diseases of the digestive system.

I manually assigned the 418 GWAS to 112 classes to aggregate identical or similar traits (Supplementary table 2).

In the EBI eQTL Catalogue, some samples have been analyzed at the cell type level and others at the tissue level.
%
For instance, immune cell types are analyzed at very high resolution that includes different stimulation of the same cell type.
%
Therefore I manually defined 35 classes to aggregate eQTL cell types into tissues (Supplementary table 1).
%
For instance, the different immune cell types such as monocyte, T-cell, etc are aggregated into an ImmuneCell class (Supplementary table 1).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Which eQTLs are the most pleiotropic?}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% TODO CONTINUE
These trait class annotations were used to investigate pleiotropy of eQTLs.
%
Different trait classes of eQTLs were aggregated and sorted according to the number of GWAS classes the eQTLs belong to (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The most pleiotropic variants belong to 6 classes that include different traits such as autoimmune diseases, circulatory disases and cancer.
%
The most pleiotropic eQTLs are found in cytobands 1p13.2, 12q24.12 and 19.q13.33.
%
The eQTL rs2476601 is well known for rheumatoid arthritis, juvenile athritis, diabetes and cancer of skin REFs.
%
Our analysis annotates this variant with eQTL genes AP4B1,AP4B1-AS1,BCL2L15,DCLRE1B,MAGI3,PHTF1 and PTPN22 tissues Adipose,Brain,Colon,Fibroblast,ImmuneCell LCL and iPSC

% TODO add rs2522051
This variant rs2522051 is involved in six classes of traits, namely,
allergy, autoimmune diseases, cancer of the breast, circulatory system diseases, hypertension, and respiratory system diseases
(Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
The variant rs2522051 is located downstream of the IRF1 gene in cytoband 5q31.1
%
This variant rs2522051 is active in 23 classes of tissues including adipose, brain, digestive cells, immune cells, and sexual organs.
%
This variant rs2522051 regulates several genes such as IL13, PDLIM4, and RAD50.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Which genomic regions contain the most pleiotropic eQTLs?}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Pleiotropic eQTLs are concentrated in genomic regions such as in cytobands 3q23,
5q31.1, 9p21.3, 15q24.1, and 19q13.33 for eQTLs involved on five or more classes (Table \ref{tab:pleitropic_variants} and Supplementary table 4).
%
Therefore, I wanted to compute pleiotropic regions with high concentration of pleiotropic eQTLs.
%
To compute these pleiotropic regions, I use a sliding window to aggregate regions with pleiotropic eQTLs at distances
of less than 1e5 nt (See Methods) (Table \ref{tab:pleiotropic_regions} and Supplementary table 5).
%
% TODO add data
I tested three SNP.H4.PP cutoffs, namely, 0.25, 0.5 and 0.75, to compute these regions.
%
% TODO
For the SNP.PP.H4 cutoff 0.75, we obtain very few regions.
%
A less stringent cutoff 0.5, gives more pleiotropic regions, but still contains the pleiotropic regions found for a
cutoff SNP.PP.H4 0.75.
%
A lower cutoff 0.25 gives more pleiotropic regions, but these regions are different than for the higher cutoffs.
%
Therefore, we fixed our analysis at SNP.PP.H4 0.5.

%
% TODO Verify values
I found seven regions with seven or more trait classes, 30?? regions with 4?? or more trait
classes and 80?? regions with 3?? or more trait classes (Supplementary table 5).
%
% TODO Verify values
50\% of regions with at least two trait classes are shorter than 100 kb, 20\% of regions are
between 100 kb and 200 kb, 10\% between 200 kb and 300 kb, and 20\% are larger than 300 kb (Fig. \ref{fig:pleiotropy_region_distribution}a).

The most pleiotropic region is 5:131,912,097-132,802,472 in cytoband 5q31.1 which contains the interferon response factor 1
(IRF1) and interleukins IL3, IL4, IL5 and IL13 (Table \ref{tab:pleiotropic_regions}).
%
Interferons and interleukins are very important factors for the immune system and anti-viral, inflammation and cancer responses.
%
The largest region is 7:2,712,518-7,254,268 in cytoband 7p22.3 with a length of 4,541,751 bp and variants associated to
autoimmune and respiratory diseases and height (Supplementary table 5).

Very pleiotropic regions with 5 trait classes or more make around 0.6 Mb of the genome,
regions with 4 GWAS classes make an additional 1 Mb and regions with 3 trait classes make an additional 1.5 Mb (Fig. \ref{fig:pleiotropy_region_distribution}b).
%
Altogether, pleiotropic regions with 3 or more trait classes make around 3 Mb of the genome (Fig. \ref{fig:pleiotropy_region_distribution}b).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{How specific are eQTLs for traits, genes and tissues?}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Next I verified whether the colocalized eQTL/GWAS variants tend to be specific to traits, eQTL genes and eQTL tissues.
%
74\% of colocalized eQTL/GWAS variants are associated to one GWAS class, 21\% are associated to 2 classes and the remaining variants associated to
3 or more trait classes make less that 4\% of colocalized eQTL/GWAS variants (Fig. \ref{fig:hist_gwas_egene_etissue}a).
%
Regarding eQTL genes, 36\% of the colocalized eQTL/GWAS variants modulate a single gene, 24\% modulate two genes, and the
remaining 40\% modulate three or more genes (Fig. \ref{fig:hist_gwas_egene_etissue}b).
%
Concerning tissues, 27\% colocalized eQTL/GWAS variants are active in a single tissue, 16\% in two tissues, 11\% in three
tissues, and the remaining 46\% in four or more tissues (Fig. \ref{fig:hist_gwas_egene_etissue}c).
%
In conclusion, trait pleiotropy is rather rare among colocalized eQTL/GWAS variants, while, regulation of at least two
genes in at least two tissues is rather common (Fig. \ref{fig:hist_gwas_egene_etissue}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Relation between trait, eQTL gene and tissue counts}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In Supplementary Figure \ref{fig:region_gwas_egenes_tissues}, I have plotted the number of trait classes, eQTL genes and tissues of
eQTLs in four pleiotropic regions.
%
All three regions are very pleiotropic with very different associated trait classes such as cancer, cardiovascular and
autoimmune diseases (Table \ref{tab:pleiotropic_regions} and Supplementary table 5).
%
However these plots show that the relationship between pleiotropy and the number of eQTL genes and tissues is not direct.
%
These plots suggest that counts of eQTL genes and tissues per variant correlate better between them than with the counts of the trait classes.
%
Indeed, Spearman correlation between eQTL gene and tissue counts is 0.75 while their correlation with the count of GWAS classes is 0.29 and 0.25, respectively (Fig. \ref{fig:region_gwas_egenes_tissues}).

In summary, counts of eQTL genes and tissues is highly correlated, while correlation with GWAS classes is much lower.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{What is the impact of pleiotropic eQTLs?}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\subsection*{What are the mechanisms of pleiotropy}

In the following sections, I studied different molecular mechanisms of pleiotropic eQTLs.
%
%I hypothesized that pleiotropy arises from bias in regulatory effects of variants.
%%
%For instance, pleiotropic variants might significantly affect some molecular functions more often.
%%
%Another possibility is that pleiotropic variants affect more eGenes in more eTissues, which affect more GWAS traits.

First, I analysed whether there are significant differences of some effect consequences between more and less pleiotropic eQTLs.
%
I separated eQTLs according to the count of GWAS classes and computed the EBI variant effect predictor (VEP) consequence (\citep{2016.Cunningham.McLaren}).
%
I found a significant larger number of upstream, 5'-UTR, downstream and 3'-UTR eQTLs
with 2, 3 and 4 and more classes compared to eQTLs with one class (Fig. \ref{fig:vep_consequence}).
%
I also found a significant larger number of intron, and non-coding transcript eQTLs
with 3 and 4 and more classes (Fig. \ref{fig:vep_consequence}).
%
These analyses suggest that more pleiotropic eQTLs have a stronger effect on
the coding sequence and splicing regions, which might explain partly their more pleiotropic function.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Do pleiotropic eQTLs modulate more genes?}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Then I hypothesized that pleiotropic eQTLs modulate more genes even after taking into account the same tissues.
%
To test this hypothesis, I computed the number of eQTL genes per eQTL-tissue pair for different trait class counts.
%
%If the GWAS class count of a variant changed in different variant-eGene-eTissue trios, then we kept the maximal one.
%
eQTLs involved in 1, 2, 3, 4, and 5 and more GWAS classes modulate an average of 1.3, 1.5, 1.6, 1.9 and 1.8 genes, respectively (Fig. \ref{fig:gwas_egene_etisue_per_variant}a).
%I found means of eGene counts of 1.4, 1.7, 1.7 and 1.6 for GWAS classes counts 5, 4, 3 and 2 compared to eGene count mean of 1.5 for GWAS class count 1 (Fig. \ref{fig:gwas_egene_etisue_per_variant}a).
%
This suggests that pleiotropic eQTLs modulate more genes compared to eQTLs involved in only one GWAS class (Fig. \ref{fig:gwas_egene_etisue_per_variant}a).
%
More target genes of eQTLs provides an explanation to the observed pleiotropy.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Are pleiotropic eQTLs active in more tissues?}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

My next hypothesis was that pleiotropic eQTL-gene pairs are active in more tissues than non-pleiotropic eQTLs.
%
To test this hypothesis, the number of tissues per eQTL-gene pair was counted for eQTL involved in different GWAS class counts.
%
eQTLs involved in 1, 2, 3, 4, and 5 and more GWAS classes are active in an average of 2.4, 2.5, 2.8, 3.4 and 4.3 tissues, respectively (Fig. \ref{fig:gwas_egene_etisue_per_variant}a).
%
This suggests that pleiotropic eQTL are active in more tissues (Mann–Whitney U test) compared to non-pleiotropic eQTLs (Fig. \ref{fig:gwas_egene_etisue_per_variant}b).

%\subsection*{Are pleiotropic eQTLs associated to more traits in unique gene-tissue pairs?}
%
%My next hypothesis was that pleiotropic variants are associated to more GWAS classes even after taking into account differences in eGene and eTissue counts.
%
%To test this hypothesis, the GWAS class counts per variant-eGene-eTissue trios were counted.
%%
%Then these GWAS categorie counts per variant-eGene-eTissue trios were classified according to the GWAS class count per variant.
%%
%If the GWAS class count per variant changed in different variant-eGene-eTissue trios, then we kept the maximal GWAS class count per variant.
%%
%I found means of GWAS class counts per variant-eGene-eTissue of 3, 2, 1.6, 1.4 and 1 for GWAS classes counts per variant of 5, 4, 3, 2 and 1 (Fig. \ref{fig:gwas_egene_etisue_per_variant}c).
%%
%This shows a significant larger number of GWAS classes even in unique trios of variant-eGene-eTissue (Fig. \ref{fig:gwas_egene_etisue_per_variant}c).
%%
%This observation could be explained by my previous observation that pleiotropic variants have more often missense effects (Fig. \ref{fig:vep_consequence}a).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection*{Are pleiotropic eQTLs bound by more transcription factors?}
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In Fig. \ref{fig:gwas_egene_etisue_per_variant}b, I observed that pairs of eQTL-gene are active in more tissues.
%
One explanation might be that pleiotropic eQTLs bound by more transcription factors, which in turn, upregulate the pleiotropic eQTLs in more tissues.
%
To test this hypothesis, I counted the number of unique transcription factors bound in a radius of 50 bp around each eQTL (Window of 100 bp).
%
eQTLs involved in 1, 2, 3 and 4 and more GWAS classes bind a mean number of 17,
19, 23 and 27 transcription factors, respectively (Fig. \ref{fig:freq_tf_per_variant}a).
%I found a significant larger number of transcription factors bound around pleiotropic variants (Fig. \ref{fig:freq_tf_per_variant}a).

Cis-regulatory modules (CRMs) are non-coding genomic regions with a higher density of bound cis-regulatory modules \citep{2021.Ballester.Hammal}.
%
I found that the odds ratio of variants annotated with CRMs vs non-annotated is significantly (Fisher's exact test) higher for variants with 2 and 3 GWAS class counts compared to class count 1 (Fig. \ref{fig:freq_tf_per_variant}b).

In summary, pleitropic eQTLs bind more unique transcription factors and are more
likely to belong to cis-regulatory modules.

%\subsection*{Do pleiotropic eQTLs have stronger effect sizes?}
%
%Pleiotropic eQTLs affect simultaneously more trait classes.
%%
%Therefore, I wonder about the relationship between pleiotropy, effect size and significance of eQTLs.
%%
%I found that the mean of the absolute eQTL effect sizes (beta) decreased between GWAS class counts 1 and 5 (Fig. \ref{fig:beta_pval}a).
%%
%Regarding the eQTL significance (Negative decimal logarithm of the p-values), I found decreasing mean values for GWAS class counts between 1 and 5 (Fig. \ref{fig:beta_pval}b).
%
%Then I carried out the same analysis for the GWAS effect size (beta) and significance (p-value).
%%
%I found that the mean of the absolute GWAS effect sizes (beta) decreased between GWAS class counts 1 and 5 (Fig. \ref{fig:beta_pval}b).
%%
%Regarding the GWAS variant significance (Negative decimal logarithm of the p-values), I found increasing mean values for GWAS class counts between 1 and 5 (Fig. \ref{fig:beta_pval}b).
%
%These observations suggest that the strength of both eQTL and GWAS effects decrease with the pleiotropy (Fig. \ref{fig:beta_pval}a,b).
